Seasonal Patterns and Peak Demand in Toronto Island Ferry Operations

### 2024-09-27 paper/paper.qmd
Fixed small typos and duplicate words

By addressing the gaps in operational planning, particularly during peak seasons, this analysis offers actionable data-driven results for ferry operators. The findings emphasize the importance of aligning ferry schedules with demand fluctuations and offer suggestions for optimizing resource allocation. The rest of the paper is structured as follows: the Data section introduces the dataset and the key variables, the Results section highlights the key findings from the time series analysis, and the Discussion provides a summary of the implications and recommendations for future research.

The remainder of the paper is structured as follows. @sec-data introduces the dataset and its source, as well as key variables used in the analysis. @sec-result presents the findings from the time series analysis, including trend detection, seasonality, and peak times of usage. @sec-disc provides a discussion of these results, highlighting key findings and offering suggestions for future work.

# Data {#sec-data}
## Data Overview {#sec-overview}
The dataset used in this analysis is the "Toronto Island Ferry Ticket Sales and Redemptions" dataset from Open Data Toronto [@citeopendatatoronto], with the time frame set to 2016-2023.The dataset provides near real-time information on ticket sales and redemptions for ferries traveling to and from Centre Island, Hanlan’s Point, and Ward’s Island. Data is recorded at 15-minute intervals and includes both point-of-sale (POS) and online ticket purchases. Toronto’s public ferries are a important transportation link for residents and visitors to the islands, and analyzing this dataset can provide valuable assessment into temporal patterns of ferry usage, helping to optimize ferry schedules and improve service [@citeopendatatoronto].

@@

### Dataset Selection
Only a few datasets related to transportation in Toronto were identified in the City of Toronto’s Open Data Catalogue, but none were as directly relevant to ferry operations as this dataset. Other datasets, such as those for city traffic or public transit, do not provide the level of detail required for a focused study on ferry operations. Therefore, the "Toronto Island Ferry Ticket Sales and Redemptions" dataset was chosen as the best fit for this analysis.

@@

### Peak Day Analysis
@tbl-table2 lists the peak ferry ticket sales for each year, highlighting specific days, often during the summer months, when ferry demand is at its highest. These peak days align with public holidays, long weekends, and special events that draw more visitors to the islands. For example, August 1st, 2016 is a Provincial Day and July 1st, 2018 is Canada Day[@citetimeanddate].In 2020, ferry sales decreased due to the pandemic, with July 25 marking a much lower peak day compared to previous years.

@@

4. **Peak Day Analysis**:
   The analysis of peak days in each year shows that significant surges in ferry ticket sales often occur around public holidays, long weekends, or special events. For instance, peak days like August 1, 2016, and August 20, 2017, coincide with well-known holiday weekends in Canada, which typically see more families and groups visiting the islands for leisure activities.

>>>>

By addressing the gaps in operational planning, particularly during peak seasons, this analysis offers actionable data-driven results for ferry operators. The findings emphasize the importance of aligning ferry schedules with demand fluctuations and offer suggestions for optimizing resource allocation. The remainder of the paper is structured as follows. @sec-overview introduces the dataset and its source, as well as key variables used in the analysis. @sec-result presents the findings from the time series analysis, including trend detection, seasonality, and peak times of usage. @sec-disc provides a discussion of these results, highlighting key findings and offering suggestions for future work.

# Data {#sec-data}
## Data Overview {#sec-overview}
The dataset used in this analysis is the "Toronto Island Ferry Ticket Sales and Redemptions" dataset from Open Data Toronto, with the time frame set to 2016-2023 [@citeopendatatoronto].The dataset provides near real-time information on ticket sales and redemptions for ferries traveling to and from Centre Island, Hanlan’s Point, and Ward’s Island. Data is recorded at 15-minute intervals and includes both point-of-sale (POS) and online ticket purchases. Toronto’s public ferries are a important transportation link for residents and visitors to the islands, and analyzing this dataset can provide valuable assessment into temporal patterns of ferry usage, helping to optimize ferry schedules and improve service [@citeopendatatoronto].

@@

### Dataset Selection
Only a few datasets related to transportation in Toronto were identified in the City of Toronto’s Open Data Catalogue, but none were as directly relevant to ferry operations as this dataset [@citeopendatatoronto]. Other datasets, such as those for city traffic or public transit, do not provide the level of detail required for a focused study on ferry operations [@citeopendatatoronto]. Therefore, the "Toronto Island Ferry Ticket Sales and Redemptions" dataset was chosen as the best fit for this analysis.

@@

### Peak Day Analysis
@tbl-table2 lists the peak ferry ticket sales for each year, highlighting specific days, often during the summer months, when ferry demand is at its highest. These peak days align with public holidays, long weekends, and special events that draw more visitors to the islands. In 2020, ferry sales decreased due to the pandemic, with July 25 marking a much lower peak day compared to previous years.

@@

4. **Peak Day Analysis**:
   The analysis of peak days in each year shows that significant surges in ferry ticket sales often occur around public holidays, long weekends, or special events. For example, August 1st, 2016 is Provincial Day and July 1st, 2018 is Canada Day, which typically see more families and groups visiting the islands for leisure activities [@citetimeanddate].
